rethinking and the 5 genera

### 2021-09-27 scripts/54_rethinking_binary.Rmd
finally donne with this yeaaaaaaaaaaaah

```{r}
library(rethinking)
library(tidyverse) 
library(ggpubr)

@@

```{r}
meta <- read_csv('../data/amplicon/stool/combined_2_meta.csv') %>% 
  mutate(logdiv_s = scale(log(simpson_reciprocal))) %>%  
  mutate(tox = if_else(toxicity == 'yes', 1, 0),
         cr100 = if_else(cr_d100 == 'yes', 1, 0),
         loca = if_else(center == 'M', 1, 2)) %>%  # MSK 1 ; Upenn 2
  mutate(center = factor(center),
         toxicity = factor(toxicity, levels = c('no','yes')),
         cr_d100 = factor(cr_d100, levels = c('no','yes'))) %>%
  # have a 2 cutpoint for the CRS
  mutate(crs3 = case_when(
    crsgrade == 0 ~ 1,
    crsgrade == 1 ~ 2,
    crsgrade >= 2 ~ 3
  ))

@@

mtox <- ulam(
    alist(
        tox ~ dbinom( 1 , p ) ,
        logit(p) <- b*logdiv_s + a[location] ,
        b ~ dnorm( 0 , 2), 
        a[location] ~ dnorm( 0 , 0.5 )
    ) , data=dat_list , chains=4 , log_lik=TRUE , cores = 8)

# the std of b is 2, since we assume the microbiome could have big impact on immunology (based on our previous understanding and knowledge )
precis(mtox,  prob = 0.95)
plot(precis(mtox, depth = 2, prob = 0.95 ) , pars = ('b'), labels = c('simpson_reciprocal'))
# there is baseline difference between penn and Msk and MSK has higher tox rage in general.
# it doesn't seem diversity has a clear association with tox rate after correcting for center wise difference
div_value <- c(-1, 1)

```{r}
# another model with the crs grade 0,1,>=2 
mcrs3 <- ulam(
    alist(
        crs3 ~ dordlogit( phi , cutpoints ),
        phi <- b*logdiv_s + a[location] ,
        b ~ dnorm( 0 , 1),  
        a[location] ~ dnorm( 0 , 0.5 ),
        cutpoints ~ dnorm( 0 , 1.5 )
    ) , data=dat_list , chains=4 , log_lik=TRUE , cores = 8)
plot(precis(mcrs3, depth = 2, prob = 0.95) ) 
# the cutpoints are not of much interet, this suggests that the diversity really isn't having a correlation?
```

@@

@@

```{r}
mcr <- ulam( 
    alist( 
        cr100 ~ dbinom( 1 , p ) ,
        logit(p) <- b*logdiv_s + a[location] ,
        b ~ dnorm( 0 , 2),  
        a[location] ~ dnorm( 0 , 0.5 )
    ) , data=dat_list , chains=4 , log_lik=TRUE , cores = 8)

@@

```{r}
# doing a prior check
prior <- extract.prior( mcr , n=1e4 )

@@

    diff = abs( cols[[1]] -cols[[2]] )
    ret = quantile(diff , probs = c(0.025, 0.5, 0.975))
    return(ret)
  })
```
### posterior

```{r}
# posterior prediction check
post <- extract.samples(mcr)
post_cr <- list(m = 1, p = 2) %>% 
  map(function(center_){
    cols = seq(1, length(div_value)) %>% 
      set_names(seq(1, length(div_value))) %>% 
      map(function(idx){
         inv_logit( post$b * div_value[idx] + post$a[, center_] ) 
      }) 
    return(cols)
  })

# plot the histogram of the probs corresponding to low and high diversity
# msk
tibble(
  prior_low = prior_cr %>% pluck('m') %>% pluck(1),
  prior_high = prior_cr %>% pluck('m') %>% pluck(2)
)%>% gather() %>% 
  gghistogram(x = 'value')+
  facet_grid(key ~ .)

>>>>

```{r}
library(rethinking)
library(tidyverse) 
library(ggpubr)
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
library(tidybayes)
```

@@

```{r}
meta <- read_csv('../data/amplicon/stool/combined_2_meta.csv') %>% 
  mutate(logdiv_s = as.numeric(scale(log(simpson_reciprocal)))) %>%  
  mutate(tox = if_else(toxicity == 'yes', 1, 0),
         cr100 = if_else(cr_d100 == 'yes', 1, 0),
         loca = if_else(center == 'M', 1, 2)) %>%  # MSK 1 ; Upenn 2
  mutate(center = factor(center),
         toxicity = factor(toxicity, levels = c('no','yes')),
         cr_d100 = factor(cr_d100, levels = c('no','yes')))

@@

mtox <- ulam(
    alist(
        tox ~ dbinom( 1 , p ) ,
        logit(p) <- b*logdiv_s + a[location] ,
        b ~ dnorm( 0 , 2), 
        a[location] ~ dnorm( 0 , 0.5 )
    ) , data=dat_list , chains=4 , log_lik=TRUE , cores = 16)

# the std of b is 2, since we assume the microbiome could have big impact on immunology (based on our previous understanding and knowledge )
precis(mtox,  prob = 0.95)
plot(precis(mtox, depth = 2, prob = 0.95 ) , pars = ('b'), labels = c('simpson_reciprocal'))

# make this plot myself with a pre-defined range in the axis
df <- precis(mtox,  prob = 0.95)  %>% as.tibble()
df <- df %>%
  mutate(variable = 'simpson_reciprocal')

#remove row number 1 (The intercept)
df %>% 
  ggplot(aes(x = reorder(variable, mean), y = mean)) +
  geom_point(shape = 15,
             size  = 4, width = 0.1,
             position = "dodge", color="black") +
  geom_errorbar(aes(ymin  = `2.5%`,
                    ymax  = `97.5%`),
                width = 0.1,
                size  = 0.7,
                position = "dodge", color="turquoise4") +
  theme(axis.title = element_text(face = "bold")) +
  xlab("Variables") + ylab("Coefficient with 95% CI") +
  coord_flip(ylim = c(-0.7, 1)) +
  #geom_hline(yintercept = 1, color = "red", size = 1) +
   theme(axis.title = element_text(size = 0)) +
  theme(axis.text = element_text(size = 14)) +
  labs(title = 'Toxicity')

ggsave('../figs/amplicon/bayesian_div_tox.pdf', width = 5, height = 4)
```
There is baseline difference between penn and Msk and MSK has higher tox rage in general.
it doesn't seem diversity has a clear association with tox rate after correcting for center wise difference

@@

    diff =  cols[[1]] -cols[[2]] 
    ret = quantile(diff , probs = c(0.025, 0.5, 0.975))
    return(diff)
  }) 
boxplot(prior_tox$p)
hist(c(prior_tox$m$lowdiv, prior_tox$p$lowdiv)) 
hist(c(prior_tox$m$highdiv, prior_tox$p$highdiv))
```

@@

```{r}
mcr <- ulam( 
    alist( 
        cr100 ~ dbinom( 1 , p ) ,
        logit(p) <- b*logdiv_s + a[location] ,
        b ~ dnorm( 0 , 2),  
        a[location] ~ dnorm( 0 , 0.5 )
    ) , data=dat_list , chains=4 , log_lik=TRUE , cores = 16)

@@

```{r}
# make this plot myself with a pre-defined range in the axis
df <- precis(mcr,  prob = 0.95)  %>% as.tibble()
df <- df %>%
  mutate(variable = 'simpson_reciprocal')

#remove row number 1 (The intercept)
df %>% 
  ggplot(aes(x = reorder(variable, mean), y = mean)) +
  geom_point(shape = 15,
             size  = 4, width = 0.1,
             position = "dodge", color="black") +
  geom_errorbar(aes(ymin  = `2.5%`,
                    ymax  = `97.5%`),
                width = 0.1,
                size  = 0.7,
                position = "dodge", color="turquoise4") +
  theme(axis.title = element_text(face = "bold")) +
  xlab("Variables") + ylab("Coefficient with 95% CI") +
  coord_flip(ylim = c(-0.7, 1)) +
  #geom_hline(yintercept = 1, color = "red", size = 1) +
   theme(axis.title = element_text(size = 0)) +
  theme(axis.text = element_text(size = 14)) +
  labs(title = 'CR') 
ggsave('../figs/amplicon/bayesian_div_cr.pdf', width = 5, height = 4)
```

```{r}
# doing a prior check
prior <- extract.prior( mcr , n=1e4, cores =16 )

@@

    diff = cols[[1]] -cols[[2]] 
    ret = quantile(diff , probs = c(0.025, 0.5, 0.975))
    return(diff)
  })

boxplot(prior_cr$p)
hist(c(prior_cr$m$lowdiv, prior_cr$p$lowdiv)) 
hist(c(prior_cr$m$highdiv, prior_cr$p$highdiv)) 
```
### posterior

```{r}
# posterior prediction check
post <- extract.samples(mcr)
post_cr <- list(m = 1, p = 2) %>% 
  map(function(center_){
    cols = seq(1, length(div_value)) %>% 
      set_names(seq(1, length(div_value))) %>% 
      map(function(idx){
         inv_logit( post$b * div_value[idx] + post$a[, center_] ) 
      }) 
    diff =  cols[[1]] -cols[[2]] 
    return(diff)
  })

hist(c(post_cr$m, post_cr$p)) 
hist(c(post_cr$m$highdiv, post_cr$p$highdiv))
```

```{r}
# if putting them together (the prior and the post)
both <- bind_rows(
  data_frame(
      coeff = c(prior_cr$m, prior_cr$p),
      grp = 'prior'
    ),
  data_frame(
      coeff = c(post_cr$m, post_cr$p),
      grp = 'post'
    )
)

both %>% 
    ggplot(aes(x = coeff, y = grp, color = grp)) +
    stat_pointinterval(.width = c(.66, .95))
```

```{r}
# plot the histogram of the probs corresponding to low and high diversity
# msk
tibble(
  prior_low = prior_cr %>% pluck('m') %>% pluck(1),
  prior_high = prior_cr %>% pluck('m') %>% pluck(2)
)%>% gather() %>% 
  gghistogram(x = 'value')+
  facet_grid(key ~ .)

### 2021-09-27 scripts/56_rethinking_genera.Rmd
finally donne with this yeaaaaaaaaaaaah

#remove row number 1 (The intercept)
df %>% 
  ggplot(aes(x = variable, y = mean)) +
  geom_point(shape = 15,
             size  = 4, width = 0.1,
             position = "dodge", color="black") +
  geom_errorbar(aes(ymin  = `2.5%`,
                    ymax  = `97.5%`),
                width = 0.1,
                size  = 0.7,
                position = "dodge", color="turquoise4") +
  theme(axis.title = element_text(face = "bold")) +
  xlab("Variables") + ylab("Coefficient with 95% CI") +
  coord_flip(ylim = c(-1.5, 1.5)) +
  geom_hline(yintercept = 0, color = "red", size = 1) +
   theme(axis.title = element_text(size = 0)) +
  theme(axis.text = element_text(size = 14)) +
  labs(title = 'Toxicity') +
  ggsave('../figs/amplicon/bayesian_genera_tox.pdf', width = 5, height = 4)
```

```{r}
post <- extract.samples(gtox) 
meta_ <- meta %>% 
  select(Akkermansia:Ruminococcus)

@@

# plot them together in one 
bind_rows(
  res_top %>% mutate(grp = 'top'),
  res_bot %>% mutate(grp = 'bot')
) %>% 
  gghistogram(x = 'value',bins = 30, fill = 'grp', palette = 'nejm', color = 'white', add = 'mean',
                  xlab = 'Predicted probability of Toxicity', ylab = 'Probability density') +
  ggsave('../figs/predicted_tox_bacteroides_top_bottom_10.pdf', width = 7, height = 5)

@@

@@

```{r}
post <- extract.samples(gtox) 
# using the patient with highest and lowest Bacteroides data to plug in
# find the patients
bl <- meta %>% 
  arrange(Bacteroides) %>% 
  slice(1) %>% 
  select(Akkermansia:Ruminococcus)

bh <- meta %>% 
  arrange(-Bacteroides) %>% 
  slice(1) %>% 
  select(Akkermansia:Ruminococcus)

pl <- inv_logit( post$ba * (-1.612806	) + post$bb * (-5.69897		) + post$be * (-5.69897	) +  post$bf * (-5.69897		) + post$br * (-0.5244935) + post$a[, 1] )

ph <- inv_logit( post$ba * (-2.286752	) + post$bb * (-0.4575058) + post$be * (-1.745601	) +  post$bf * (-5.69897) + post$br * (-4.050354) + post$a[, 1] )

bind_rows(
  tibble(
    value = pl,
    grp = 'low'
  ),
  tibble( 
    value = ph,
    grp = 'high'
  )
) %>% 
  gghistogram(x = 'value',bins = 30, fill = 'grp', palette = 'nejm', color = 'white',add = 'mean',title = 'MSK',
              xlab = 'Predicted probability of toxicity', ylab = 'Probability density')

bind_rows(
  tibble(value = pl,grp = 'low'),
  tibble( value = ph,grp = 'high')
)   %>% 
  group_by(grp) %>% 
  summarise(m = mean(value))
```

```{r}
gcr <- ulam(
    alist(
        cr100 ~ dbinom( 1 , p ) ,
        logit(p) <- ba*Akkermansia + bb*Bacteroides + be*Enterococcus + bf*Faecalibacterium + br*Ruminococcus + a[location] ,
        ba ~ dnorm( 0 , 1),
        bb ~ dnorm( 0 , 1),
        be ~ dnorm( 0 , 1),
        bf ~ dnorm( 0 , 1),
        br ~ dnorm( 0 , 1),
        a[location] ~ dnorm( 0 , 0.5 )
    ) , data=dat_list , chains=4 , log_lik=TRUE , cores = 8)
precis(gcr, prob = 0.95) 
plot(precis(gcr, depth = 2, prob = 0.95) ) 
# I think the Ruminococcus still holds for the CR
```

#remove row number 1 (The intercept)
df %>% 
  ggplot(aes(x = variable, y = mean)) +
  geom_point(shape = 15,
             size  = 4, width = 0.1,
             position = "dodge", color="black") +
  geom_errorbar(aes(ymin  = `2.5%`,
                    ymax  = `97.5%`),
                width = 0.1,
                size  = 0.7,
                position = "dodge", color="turquoise4") +
  theme(axis.title = element_text(face = "bold")) +
  xlab("Variables") + ylab("Coefficient with 95% CI") +
  coord_flip(ylim = c(-1.5, 1.5)) +
  geom_hline(yintercept = 0, color = "red", size = 1) +
   theme(axis.title = element_text(size = 0)) +
  theme(axis.text = element_text(size = 14)) +
  labs(title = 'CR') +
  ggsave('../figs/amplicon/bayesian_genera_cr.pdf', width = 5, height = 4)
```

```{r}
post <- extract.samples(gcr) 
# using the patient with highest and lowest Ruminococcus data to plug in
# find the patients
rl <- meta %>% 
  arrange(Ruminococcus) %>% 
  slice(1) %>% 
  select(Akkermansia:Ruminococcus)

rh <- meta %>% 
  arrange(-Ruminococcus) %>% 
  slice(1) %>% 
  select(Akkermansia:Ruminococcus)

pl <- inv_logit( post$ba * (-3.517128) + post$bb * (-2.447201	) + post$be * (-2.504205	) +  post$bf * (-5.69897	) + post$br * (-5.69897) + post$a[, 1] )

ph <- inv_logit( post$ba * (-1.845325	) + post$bb * (-3.052386) + post$be * (-5.69897) +  post$bf * (-5.69897) + post$br * (-0.373353) + post$a[, 1] )

bind_rows(
  tibble(value = pl,grp = 'low'),
  tibble( value = ph,grp = 'high')
) %>% 
  gghistogram(x = 'value',bins = 30, fill = 'grp', palette = 'nejm', color = 'white',title = 'MSK', add = 'mean',
              xlab = 'Predicted probability of CR d100', ylab = 'Probability density')

bind_rows(
  tibble(value = pl,grp = 'low'),
  tibble( value = ph,grp = 'high')
)   %>% 
  group_by(grp) %>% 
  summarise(m = mean(value))
```

@@

```{r}
# plot them together in one 
bind_rows(
  res_top %>% mutate(grp = 'top'),
  res_bot %>% mutate(grp = 'bot')
) %>% 
  gghistogram(x = 'value',bins = 30, fill = 'grp', palette = 'nejm', color = 'white', add = 'mean',
                  xlab = 'Predicted probability of CR d100', ylab = 'Probability density')+
  ggsave('../figs/predicted_CR_Ruminococcus_top_bottom_10.pdf', width = 7, height = 5)

>>>>

#remove row number 1 (The intercept)
df %>% 
  ggplot(aes(x = variable, y = mean)) +
  geom_point(shape = 15,
             size  = 4, width = 0.1,
             position = "dodge", color="black") +
  geom_errorbar(aes(ymin  = `2.5%`,
                    ymax  = `97.5%`),
                width = 0.1,
                size  = 0.7,
                position = "dodge", color="turquoise4") +
  theme(axis.title = element_text(face = "bold")) +
  xlab("Variables") + ylab("Coefficient with 95% CI") +
  coord_flip(ylim = c(-1.5, 1.5)) +
  geom_hline(yintercept = 0, color = "red", size = 1) +
   theme(axis.title = element_text(size = 0)) +
  theme(axis.text = element_text(size = 14)) +
  labs(title = 'Toxicity') 
ggsave('../figs/amplicon/bayesian_genera_tox_log10.pdf', width = 5, height = 4)
```

### Prior check

I will just do something similar to the nutrition project : the prior annd post coeff forest plot

```{r}
theme_set(theme_tidybayes() + cowplot::panel_border())
# logit(p) <- ba*Akkermansia + bb*Bacteroides + be*Enterococcus + bf*Faecalibacterium + br*Ruminococcus + a[location]
prior <- extract.prior( gtox , n=4000 ) %>% as.tibble()  %>% select(-a) %>% mutate(grp = 'prior')
post <- extract.samples(gtox) %>% as.tibble()  %>% select(-a) %>% mutate(grp = 'post')
# comparing prior and post 
tox <- bind_rows(prior, post) %>% 
  gather('term', 'coeff', ba:br) %>% 
  mutate(genus = case_when(
    term == 'ba' ~ 'Akkermansia',
    term == 'bb' ~ 'Bacteroides',
    term == 'be' ~ 'Enterococcus',
    term == 'bf' ~ 'Faecalibacterium',
    term == 'br' ~ 'Ruminococcus'
  ))

post_order <- tox %>% 
  filter(grp == 'post') %>% 
  group_by(term, genus) %>% 
  summarise(q50 = median(coeff)) %>% 
  arrange(-q50) %>% 
  pull(genus)

tox %>%
  mutate(genus = factor(genus, levels = post_order)) %>% 
  ggplot(aes(x = coeff, y = grp, color = grp)) +
  stat_pointinterval(.width = c(.66, .95)) +
  scale_color_manual(values = c('#EC0000','#00468B')) +
  geom_vline(xintercept = 0, col = 'gray', linetype = 'dashed') +
  facet_grid(genus ~ .) +
  labs(x = 'Regression coefficients',
       y = '') +
  theme(legend.position = 'none')
```

```{r}

meta_ <- meta %>% 
  select(Akkermansia:Ruminococcus)

@@

# plot them together in one 
bacteroides_top_bottom_10 <- bind_rows(
  res_top %>% mutate(grp = 'top'),
  res_bot %>% mutate(grp = 'bot')
) %>% 
  gghistogram(x = 'value',bins = 30, fill = 'grp', palette = 'nejm', color = 'white', add = 'mean',
                  xlab = 'Predicted probability of Toxicity', ylab = 'Probability density')

ggsave('../figs/predicted_tox_bacteroides_top_bottom_10.pdf', width = 7, height = 5)

@@

```{r}
gcr <- ulam(
    alist(
        cr100 ~ dbinom( 1 , p ) ,
        logit(p) <- ba*Akkermansia + bb*Bacteroides + be*Enterococcus + bf*Faecalibacterium + br*Ruminococcus + a[location] ,
        ba ~ dnorm( 0 , 1),
        bb ~ dnorm( 0 , 1),
        be ~ dnorm( 0 , 1),
        bf ~ dnorm( 0 , 1),
        br ~ dnorm( 0 , 1),
        a[location] ~ dnorm( 0 , 0.5 )
    ) , data=dat_list , chains=4 , log_lik=TRUE , cores = 16)
precis(gcr, prob = 0.95) 
plot(precis(gcr, depth = 2, prob = 0.95) ) 
# I think the Ruminococcus still holds for the CR
```

#remove row number 1 (The intercept)
df %>% 
  ggplot(aes(x = variable, y = mean)) +
  geom_point(shape = 15,
             size  = 4, width = 0.1,
             position = "dodge", color="black") +
  geom_errorbar(aes(ymin  = `2.5%`,
                    ymax  = `97.5%`),
                width = 0.1,
                size  = 0.7,
                position = "dodge", color="turquoise4") +
  theme(axis.title = element_text(face = "bold")) +
  xlab("Variables") + ylab("Coefficient with 95% CI") +
  coord_flip(ylim = c(-1.5, 1.5)) +
  geom_hline(yintercept = 0, color = "red", size = 1) +
   theme(axis.title = element_text(size = 0)) +
  theme(axis.text = element_text(size = 14)) +
  labs(title = 'CR') 
ggsave('../figs/amplicon/bayesian_genera_cr_log10.pdf', width = 5, height = 4)
```

@@

@@

```{r}
# plot them together in one 
bind_rows(
  res_top %>% mutate(grp = 'top'),
  res_bot %>% mutate(grp = 'bot')
) %>% 
  gghistogram(x = 'value',bins = 30, fill = 'grp', palette = 'nejm', color = 'white', add = 'mean',
                  xlab = 'Predicted probability of CR d100', ylab = 'Probability density')
  ggsave('../figs/predicted_CR_Ruminococcus_top_bottom_10.pdf', width = 7, height = 5)
